\textcolor{black}{In-silico molecular transport via perivascular networks in the human intracranial space}

### 2025-06-27 paper/main.tex
change molecular enrichment with mol transport or tracer enrichment

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs in the SAS. A key observation is that the balance between CSF
production and intracranial pulsatility significantly shapes the
large-scale features of tracer enrichment, with e.g.~ventricular
tracer reflux after intrathecal injection in the absence of CSF
production, and preference towards anterior brain regions with reduced
pulsatility. Indeed, the simulated distribution patterns cover a
substantial range of the individual variations observed clinically
\cite{ringstad2018brain}. In terms of perivascular pathways, we find
that even moderate CSF flow on the order of 0.1--1.0 \textmu m/s in
the surface PVSs results in earlier tracer enrichment around major
cerebral arteries. This effect is more pronounced -- with up to a
three-hour difference in time of arrival -- with mean PVS flow rates
of approximately 10 \textmu m/s. Our models of CSF flow in
perivascular networks, based on first principles and asymptotic
analysis, predict net flow of such magnitudes and both antegrade and
retrograde flow within the perivascular network. The PVS may thus
function as highways facilitating rapid transport, even in the absence
of structural barriers, though a sharp concentration difference
between the PVS and surrounding tissues is unlikely without such
barriers. Finally, dilated PVS in the SAS will cause a substantial reduction in
cardiac- and vasomotion-driven flow velocities, which in turn impedes
perivascular transport.

@@

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated tracer enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion 
factor is a simplification that does not capture the full complexity
of transient fluid dynamics, such as non-uniform wall motion and ependymal
cilia within the ventricular system~\cite{siyahhan2014flow}, dynamic CSF
flow at the craniocervical junction, or the effect of microanatomical
features such as subarachnoid trabeculae on flow resistance and dispersion
in the SAS. However, our approach is supported by recent clinical 
observations. High-resolution MRI measurements by Hirschler et
al.~\cite{hirschler2024region} revealed that CSF mobility in regions
with high pulsatility is approximately tenfold greater than the 
self-diffusion of water, an enhancement factor consistent with 
the values computed in our model.

>>>>

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
transport and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs in the SAS. A key observation is that the balance between CSF
production and intracranial pulsatility significantly shapes the
large-scale features of tracer enrichment, with e.g.~ventricular
tracer reflux after intrathecal injection in the absence of CSF
production, and preference towards anterior brain regions with reduced
pulsatility. Indeed, the simulated distribution patterns cover a
substantial range of the individual variations observed clinically
\cite{ringstad2018brain}. In terms of perivascular pathways, we find
that even moderate CSF flow on the order of 0.1--1.0 \textmu m/s in
the surface PVSs results in earlier tracer enrichment around major
cerebral arteries. This effect is more pronounced -- with up to a
three-hour difference in time of arrival -- with mean PVS flow rates
of approximately 10 \textmu m/s. Our models of CSF flow in
perivascular networks, based on first principles and asymptotic
analysis, predict net flow of such magnitudes and both antegrade and
retrograde flow within the perivascular network. The PVS may thus
function as highways facilitating rapid transport, even in the absence
of structural barriers, though a sharp concentration difference
between the PVS and surrounding tissues is unlikely without such
barriers. Finally, dilated PVS in the SAS will cause a substantial reduction in
cardiac- and vasomotion-driven flow velocities, which in turn impedes
perivascular transport.

@@

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
transport and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated tracer enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
transport patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion 
factor is a simplification that does not capture the full complexity
of transient fluid dynamics, such as non-uniform wall motion and ependymal
cilia within the ventricular system~\cite{siyahhan2014flow}, dynamic CSF
flow at the craniocervical junction, or the effect of microanatomical
features such as subarachnoid trabeculae on flow resistance and dispersion
in the SAS. However, our approach is supported by recent clinical 
observations. High-resolution MRI measurements by Hirschler et
al.~\cite{hirschler2024region} revealed that CSF mobility in regions
with high pulsatility is approximately tenfold greater than the 
self-diffusion of water, an enhancement factor consistent with 
the values computed in our model.
